String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-08-02 2023-1-19/sections/algorithms.tex
grammerly check

The goal of this section is to show how we solve RECL constraints by utilizing CEFA. 
At first, we reduce the satisfiability of RECL constraints into a decision problem for CEFA defined in the sequel. Then we propose an algorithm to solve the decision problem.

@@

Let $e$ be a regular expression over $\Sigma$. Then an occurrence of counting operators in $e$, say $(e')^{\{m,n\}}$ (or $(e')^{\{m, \infty\}}$), is said to be \emph{register-representable} if $(e')^{\{m,n\}}$ (or $(e')^{\{m, \infty\}}$) is not in the scope of a Kleene star, another counting operator, complement, or language difference in $e$. We say that $e$ is \emph{register-representable} if all occurrences of counting operators in $e$ are register-representable. For instance, $(a^{\{2, 6\}} \concat b^*) \cap (a + b)^{\{4, \infty\}}$ is register-representable, while $(a^{\{2, 6\}} \concat b^*) \setminus (a + b)^{\{4, \infty\}}$ and $(a^{\{2, 6\}} \concat b^*)^{\{2,\infty\}}$ are not, since $a^{\{2,6\}}$ is in the scope of language difference and the counter operator $\{2, \infty\}$ respectively.

Let $e$ be a register-representable regular expression over $\Sigma$. We are going to construct a CEFA out of $e$, denoted by $\aut_e$, by the following procedure. 
\begin{enumerate}
\item For each maximal sub-expression $e'$ of $e$ such that $e'$ contains no occurrences of counting operators, an NFA $\aut_{e'}$ can be constructed by a structural induction on the syntax of $e'$. 
%
\item For each sub-expression $(e')^{\{m,n\}}$ with $m \le n$ (resp. $(e')^{\{m, \infty\}}$) of $e$, we construct a CEFA $\aut_{(e')^{\{m,n\}}}$ (resp. $\aut_{(e')^{\{m, \infty\}}}$) as follows. Let $\aut_{e'} = (Q, \Sigma, \delta, I, F)$. From the construction for Kleene star in Section~\ref{sec:pre}, $(\aut_{e'})^* = (Q', \Sigma, \delta', I', F')$, where $Q' = Q \cup \{q_0\}$ with $q_0 \not \in Q$, $I' = \{q_0\}$, $F'= F \cup \{q_0\}$, $\delta' = \delta  \cup \{(q_0, a, q')  \mid \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\} \cup \{(q, a, q') \mid q \in F, \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\}$. Then $\aut_{(e')^{\{m,n\}}} = ((r'), Q', \Sigma, \delta'', I', F', \alpha')$, where $r'$ is a register, 
%
$$
\begin{array}{l c l}
\delta'' & = & \{(q, a, q', (0)) \mid (q, a, q') \in \delta\}\  \cup \\
& & \{(q_0, a, q', (1))  \mid \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\}\ \cup \\
& & \{(q, a, q', (1)) \mid q \in F, \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\},
\end{array}
$$ 
%
moreover, $\alpha' = m \le r' \le n$ if $I \cap F = \emptyset$, and $\alpha' = r' \le n$ otherwise.  (Intuitively, if $\varepsilon$ is accepted by $\aut_{e'}$, then the value of $r'$ can be less than $m$.) Finally, $\aut_{(e')^{\{m, \infty\}}}$ is constructed by adapting $\alpha'$ in $\aut_{(e')^{\{m,n\}}}$ as follows: $\alpha' = m \le r'$ if $I \cap F = \emptyset$ and $\alpha' = \ltrue$ otherwise.

@@

For non-register-representable regular expressions, we first transform them into register-representable regular expressions by unfolding all the non-register-representable occurrences of  counting operators, then utilize the aforementioned procedure to construct CEFA. For instance, $(a^{\{2, 6\}} \concat b^*)^{\{2,\infty\}}$ is transformed into $(aa(\varepsilon + a + aa + aaa+aaaa) \concat b^*)^{\{2, \infty\}}$. 
The unfoldings of the inner counting operators of non-register-representable regexes incur an exponential blow-up in the worst case. Nevertheless, those regexes occupy only 5\% of the 48,834 regexes that are collected from the practice (see Section~\ref{sec:bench}). Moreover, the unfoldings are partial in the sense that the outmost counting operators are not unfolded. It turns out that our approach can solve almost all of them, except 145.   
%$(a^{\{2, 6\}} \concat b^*) \setminus (a + b)^{\{4, \infty\}}$ is transformed into $(aa(\varepsilon + a + aa + aaa + aaaa) \concat b^*) \setminus (a+b)(a+b)(a+b)(a+b)(a+b)^*$.

@@

% We must syntactically rewrite bounded repetition $\regex\{m,n\}$ if it is the sub-regex of complement (e.g., $(\regex\{m,n\})^C$), closure(e.g., $(\regex\{m,n\})^*$), and bounded repetition(e.g., $(\regex\{m,n\})\{m', n'\}$), we unwind it to $\regex^m\mid\cdots\mid\regex^n$. After this syntactic rewriting, we call the resulting regex $\regex'$ \emph{non-nested}.
Given a regex $\regex$ without an unhandled counting operator, the base case is constructing the CEFA of a single character, an empty string, or an empty language. The inductive step is constructing CEFA of the concatenation, union, intersection, closure, complement, and counting of CEFA(s) of sub-regex(es). Operations except for counting are trivial and represented in Appendix \ref{appendix:cefa}, we only illustrate counting here.

@@

      \path[->] (q_0) edge [loop below] node{$\scriptstyle\Sigma:(1)$} ();
    \end{tikzpicture}
  \end{figure}
\end{example} 
% \subsection{Combine Regex-counting and String-length}
% In the previous section, we discussed the construction of the regex-counting operator and string-length constraint. In this section, we show how to combine them to deal with the string constraints with regex-counting and string-length. 
% \begin{definition} [The Product of Two CEFA]

@@

In the sequel, we reduce the nonemptiness of CEFA in $\Lambda'_{x_1}, \cdots, \Lambda'_{x_n}$ w.r.t. $\varphi$ to the satisfiability of an LIA formula. The reduction relies on two observations.

The first observation follows from the fact that $\Lambda'_{x_i}$ for each $i \in [n]$ contains exactly one CEFA.

@@

\paragraph*{Step 2.5.} Initially, let $len := 0$, and for each $i \in [n]$, let 
%$\confset_{\cC_i}:=\emptyset$, 
$\confset_{\cC_i}:=\{(q_{i,0}, \myvec{0})\}$ 
and $\psi_{\cC_i}:= \lfalse$.  Then iterate the following procedure, until $len = B+1$.
\begin{enumerate}
\item Let $\xi: = \bigwedge \limits_{i \in [n]} \alpha_i \wedge \varphi$.
%
\item For each $i \in [n]$, do the following.
\begin{enumerate}
\item Let 
%$\confset_{\cC_i}:= \confset_{\cC_i} \cup \confset'_{\cC_i}$, 
$old_i:=\confset_{\cC_i}$, $\confset_{\cC_i}:=\emptyset$.
%
\item For each $(q, \myvec{v}) \in old_i$,
\begin{enumerate}
\item if $q \in F'_i$, then let $\psi_{\cC_i}:=\psi_{\cC_i} \vee (\bigwedge \limits_{j \in [k_i]} r_{i,j} = v_j)$, 
%
%\item let $\confset'_{\cC_i}:=\confset'_{\cC_i} \setminus \{(q, \myvec{v})\}$, 
%
\item for each $(q, \myvec{v'}, q') \in \delta'_i$, let $\confset_{\cC_i}:=\confset_{\cC_i} \cup \{(q', \myvec{v}+ \myvec{v'})\}$. 
\end{enumerate}
%
%\item 
%
\item Let $\xi := \xi \wedge \psi_{\cC_i}$. 
\end{enumerate}
%
\item If $\xi$ is satisfiable, then report $\ltrue$ and the whole algorithm terminates. 
%
\item Let $len: = len+1$. \qed
\end{enumerate}
Intuitively, for each $i \in [n]$, $\psi_{\cC_i}$ is a global variable whose value is never reset and it stores the set of ``final'' valuations of registers of $\cC_i$ (that is, the valuations obtained when reaching final states), and $\confset_{\cC_i}$ stores the set of newly discovered (not necessarily final) ``configurations'' of $\cC_i$.
%, and $\confset_{\cC_i}$ stores those that have been discovered before. 
Note that the solution searching algorithm in Step 2.5 is incremental in the sense that when the value of $len$ is incremented, the searching continues from $\confset_{\cC_i}$'s, not from scratch.
%, whose values are not reset.
%Furthermore, for each $i \in [n]$, we can remove the redundancy of characters/vectors in $\cC_i$.

%%%%%%%%%%%% the original texts by Denghang %%%%%%%%%
%%%%%%%%%%%% the original texts by Denghang %%%%%%%%%
\hide{
Moreover, the transitions $(q, \myvec{0}, q')$ in $\cC_i$ can be seen as $\varepsilon$-transitions, that is, we can replace each transition $(q, \myvec{0}, q')$ with $(q, \varepsilon, q')$, without affecting the set $\{ \sum\limits_{\myvec{v} \in V_i} m_{\myvec{v}} \myvec{v} \mid \myvec{m} \in parikh(\cC_i)\}$. Therefore, we first do these replacements, apply the $\varepsilon$-transition removal procedure. Then we utilize the determinization and minimization algorithm for NFA to reduce the size of $\cC_i$.

It turns out that to solve the nonemptiness of CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ w.r.t. $\varphi$, we can ignore the characters in $\cB_1,\cdots, \cB_n$. This is due to the fact that there is only one CEFA for each string variable and $\varphi$ only cares about the values of registers.

@@

Moreover, the transitions $(q, \myvec{0}, q')$ in $\cC_i$ can be seen as $\varepsilon$-transitions, that is, we can replace each transition $(q, \myvec{0}, q')$ with $(q, \varepsilon, q')$, without affecting the set $\{ \sum\limits_{\myvec{v} \in V_i} m_{\myvec{v}} \myvec{v} \mid \myvec{m} \in parikh(\cC_i)\}$. Therefore, we first do these replacements, apply the $\varepsilon$-transition removal procedure. Then we utilize the determinization and minimization algorithm for NFA to reduce the size of $\cC_i$.

@@

In summary, we obtain an algorithm for solving the nonemptiness of $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ w.r.t. $\varphi$, as illustrated in Figure~\ref{xxx}.

@@

As shown in the Algorithm \ref{alg:underApprox}, $\varphi_{under}$ is the under-approximated LIA formula which is initialized to the input LIA formula $\psi$ at line 2. $\varphi_\aut$ is the under-approximated LIA formula of each automaton $\aut$ which is initialized to $false$ at line 3. $lastStates$ maps each automaton $\aut$ to the set consisting of pairs of state and corresponding registers' value in the last enumeration. $lastStates(\aut)$ is initialized to the pair of the initial state and zero values for each automaton $\aut$ at line 4. $length$ is the current bound of lengths of string variables which is initialized to $0$ at line 5. The main loop at lines 6-20 iteratively tries possible solutions from low to high length bound. For each iteration, we generate the LIA formula respectively for each automaton. To each CEFA $\aut$, we get pairs of last state and corresponding registers' values from $lastStates$ at line 8. In each pair, if the last state is accepted at line 9, we add the LIA formula of the possible registers' values to $\varphi_\aut$ as a conjunct at line 10. After that, we update the set of pairs in $lastStates$ by removing the probed pair at line 12 and adding new pairs of the next state and corresponding registers' values at line 13. Finally, we combine the incremental LIA formula of each automaton at line 15. If the LIA formula $\varphi\wedge \theta$ is satisfiable at line 17, i.e., we find a solution with all strings having length less than or equal to $length$, we return \emph{true} at line 18. Otherwise, the next iteration would be invoked with a larger length. If we can not find a solution within the max string length bound, we return \emph{false} at line 22.

@@

  Consider the automaton $\aut$ in Fig. \ref{fig:overview:product:reduced} again without the LIA constraint $r_0 > 120$, i.e., $\theta$ changes to $1\leq r_1\leq 60\wedge 1\leq r_2 \leq 60\wedge 0\leq r_3 \leq 60 $. The solution of the Parikh image of it leads to a complete but inefficient result since the quantifier-free LIA formulas are solved in exponential time. Under-approximation can accelerate the solving process. In the first iteration of under-approximation, the pairs in the $lastStates(\aut)$ is the initial pair $(q_0', (0,0,0,0))$ and $\varphi_\aut$ maintains false at line 9 because $q_0'$ is not accepted. Then we update $lastStates(\aut)$ by removing the pair $(q_0', (0,0,0,0))$ and adding pairs $(q_0', (1,1,0,0)), (q_1', (1,0,1,0)), (q_2', (1,0,0,0))$. In the second iteration, $\varphi_\aut$ changes to $r_0=1\vee r_1=0\vee r_2=0\vee r_3 = 0$ because $q_2'$ is accepted. $\varphi_{under}$ is always equal to $\varphi_\aut$ since there is only one automaton. We solve the formula $\theta\wedge \varphi_{under}$ and find it is unsatisfiable. In the third iteration, the pairs in the $lastStates(\aut)$ with accepting state should be $(q_0', (2,2, 0, 0)), (q_1', (2,1,1,0))$, $(q_1', (2,0,2,0))$, $(q_2', (2,1,0,0)), (q_2', (2,0,1,0)),$ $(q_2', (2,0, 0, 0))$. The corresponding registers' values for the accepted state $q_2'$ is $(2,1,0,0), (2, 0, 1, 0), (2, 0, 0, 0)$, in which the value of $r_1$ and $r_2$ can not be both larger than 1 which is required by $\theta$. So the formula $\theta\wedge \varphi_{under}$ is still unsatisfiable. Finally, We obtain a pair $(q_2', (3,1,1,0))$ from the pair $(q_1', (2,1,1,0))$ and the transition $q_1'\xrightarrow[(1,0,0,0)]{} q_2'$, whose corresponding registers' values is a solution $(3,1,1,0)$, i.e., $r_0=3, r_1=1, r_2=1, r_3=0$.

>>>>

This section aims to show how we solve RECL constraints by utilizing CEFA. 
At first, we reduce the satisfiability of RECL constraints into a decision problem for CEFA defined in the sequel. Then we propose an algorithm to solve the decision problem.

@@

Let $e$ be a regular expression over $\Sigma$. Then an occurrence of counting operators in $e$, say $(e')^{\{m,n\}}$ (or $(e')^{\{m, \infty\}}$), is said to be \emph{register-representable} if $(e')^{\{m,n\}}$ (or $(e')^{\{m, \infty\}}$) is not in the scope of a Kleene star, another counting operator, complement, or language difference in $e$. We say that $e$ is \emph{register-representable} if all occurrences of counting operators in $e$ are register-representable. For instance, $(a^{\{2, 6\}} \concat b^*) \cap (a + b)^{\{4, \infty\}}$ is register-representable, while $(a^{\{2, 6\}} \concat b^*) \setminus (a + b)^{\{4, \infty\}}$ and $(a^{\{2, 6\}} \concat b^*)^{\{2,\infty\}}$ are not since $a^{\{2,6\}}$ is in the scope of language difference and the counter operator $\{2, \infty\}$ respectively.

Let $e$ be a register-representable regular expression over $\Sigma$. By the following procedure, we will construct a CEFA out of $e$, denoted by $\aut_e$. 
\begin{enumerate}
\item For each maximal sub-expression $e'$ of $e$ such that $e'$ contains no occurrences of counting operators, an NFA $\aut_{e'}$ can be constructed by structural induction on the syntax of $e'$. 
%
\item For each sub-expression $(e')^{\{m,n\}}$ with $m \le n$ (resp. $(e')^{\{m, \infty\}}$) of $e$, we construct a CEFA $\aut_{(e')^{\{m,n\}}}$ (resp. $\aut_{(e')^{\{m, \infty\}}}$) as follows. Let $\aut_{e'} = (Q, \Sigma, \delta, I, F)$. From the construction for Kleene star in Section~\ref{sec:pre}, $(\aut_{e'})^* = (Q', \Sigma, \delta', I', F')$, where $Q' = Q \cup \{q_0\}$ with $q_0 \not \in Q$, $I' = \{q_0\}$, $F'= F \cup \{q_0\}$, $\delta' = \delta  \cup \{(q_0, a, q')  \mid \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\} \cup \{(q, a, q') \mid q \in F, \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\}$. Then $\aut_{(e')^{\{m,n\}}} = ((r'), Q', \Sigma, \delta'', I', F', \alpha')$, where $r'$ is a register, 
%
$$
\begin{array}{l c l}
\delta'' & = & \{(q, a, q', (0)) \mid (q, a, q') \in \delta\}\  \cup \\
& & \{(q_0, a, q', (1))  \mid \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\}\ \cup \\
& & \{(q, a, q', (1)) \mid q \in F, \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\},
\end{array}
$$ 
%
moreover, $\alpha' = m \le r' \le n$ if $I \cap F = \emptyset$, and $\alpha' = r' \le n$ otherwise.  (Intuitively, if $\varepsilon$ is accepted by $\aut_{e'}$, then the value of $r'$ can be less than $m$.) Finally, $\aut_{(e')^{\{m, \infty\}}}$ is constructed by adapting $\alpha'$ in $\aut_{(e')^{\{m,n\}}}$ as follows: $\alpha' = m \le r'$ if $I \cap F = \emptyset$ and $\alpha' = \ltrue$ otherwise.

@@

For non-register-representable regular expressions, we first transform them into register-representable regular expressions by unfolding all the non-register-representable occurrences of counting operators, then utilize the aforementioned procedure to construct CEFA. For instance, $(a^{\{2, 6\}} \concat b^*)^{\{2,\infty\}}$ is transformed into $(aa(\varepsilon + a + aa + aaa+aaaa) \concat b^*)^{\{2, \infty\}}$. 
The unfoldings of the inner counting operators of non-register-representable regexes incur an exponential blow-up in the worst case. Nevertheless, those regexes occupy only 5\% of the 48,834 regexes that are collected from the practice (see Section~\ref{sec:bench}). Moreover, the unfoldings are partial in the sense that the outmost counting operators are not unfolded. It turns out that our approach can solve almost all of them except 145.   
%$(a^{\{2, 6\}} \concat b^*) \setminus (a + b)^{\{4, \infty\}}$ is transformed into $(aa(\varepsilon + a + aa + aaa + aaaa) \concat b^*) \setminus (a+b)(a+b)(a+b)(a+b)(a+b)^*$.

@@

% We must syntactically rewrite bounded repetition $\regex\{m,n\}$ if it is the sub-regex of complement (e.g., $(\regex\{m,n\})^C$), closure(e.g., $(\regex\{m,n\})^*$), and bounded repetition(e.g., $(\regex\{m,n\})\{m', n'\}$), we unwind it to $\regex^m\mid\cdots\mid\regex^n$. After this syntactic rewriting, we call the resulting regex $\regex'$ \emph{non-nested}.
Given a regex $\regex$ without an unhandled counting operator, the base case is constructing the CEFA of a single character, an empty string, or an empty language. The inductive step is constructing CEFA of the concatenation, union, intersection, closure, complement, and counting of CEFA(s) of sub-regex(es). Operations except for counting are trivial and represented in Appendix \ref{appendix:cefa}; we only illustrate counting here.

@@

      \path[->] (q_0) edge [loop below] node{$\scriptstyle\Sigma:(1)$} ();
    \end{tikzpicture}
  \end{figure}
\end{example} 
% \subsection{Combine Regex-counting and String-length}
% In the previous section, we discussed the construction of the regex-counting operator and string-length constraint. In this section, we show how to combine them. 
% \begin{definition} [The Product of Two CEFA]

@@

In the sequel, we reduce the non-emptiness of CEFA in $\Lambda'_{x_1}, \cdots, \Lambda'_{x_n}$ w.r.t. $\varphi$ to the satisfiability of an LIA formula. The reduction relies on two observations.

The first observation follows that $\Lambda'_{x_i}$ for each $i \in [n]$ contains exactly one CEFA.

@@

\paragraph*{Step 2.5.} Initially, let $len := 0$, and for each $i \in [n]$, let 
%$\confset_{\cC_i}:=\emptyset$, 
$\confset_{\cC_i}:=\{(q_{i,0}, \myvec{0})\}$ 
and $\psi_{\cC_i}:= \lfalse$.  Then iterate the following procedure until $len = B+1$.
\begin{enumerate}
\item Let $\xi: = \bigwedge \limits_{i \in [n]} \alpha_i \wedge \varphi$.
%
\item For each $i \in [n]$, do the following.
\begin{enumerate}
\item Let 
%$\confset_{\cC_i}:= \confset_{\cC_i} \cup \confset'_{\cC_i}$, 
$old_i:=\confset_{\cC_i}$, $\confset_{\cC_i}:=\emptyset$.
%
\item For each $(q, \myvec{v}) \in old_i$,
\begin{enumerate}
\item if $q \in F'_i$, then let $\psi_{\cC_i}:=\psi_{\cC_i} \vee (\bigwedge \limits_{j \in [k_i]} r_{i,j} = v_j)$, 
%
%\item let $\confset'_{\cC_i}:=\confset'_{\cC_i} \setminus \{(q, \myvec{v})\}$, 
%
\item for each $(q, \myvec{v'}, q') \in \delta'_i$, let $\confset_{\cC_i}:=\confset_{\cC_i} \cup \{(q', \myvec{v}+ \myvec{v'})\}$. 
\end{enumerate}
%
%\item 
%
\item Let $\xi := \xi \wedge \psi_{\cC_i}$. 
\end{enumerate}
%
\item If $\xi$ is satisfiable, then report $\ltrue$ and the whole algorithm terminates. 
%
\item Let $len: = len+1$. \qed
\end{enumerate}
Intuitively, for each $i \in [n]$, $\psi_{\cC_i}$ is a global variable whose value is never reset, and it stores the set of ``final'' valuations of registers of $\cC_i$ (that is, the valuations obtained when reaching final states), and $\confset_{\cC_i}$ stores the set of newly discovered (not necessarily final) ``configurations'' of $\cC_i$.
%, and $\confset_{\cC_i}$ stores those that have been discovered before. 
Note that the solution searching algorithm in Step 2.5 is incremental because when the value of $len$ is incremented, the searching continues from $\confset_{\cC_i}$'s, not from scratch.
%, whose values are not reset.
%Furthermore, for each $i \in [n]$, we can remove the redundancy of characters/vectors in $\cC_i$.

%%%%%%%%%%%% the original texts by Denghang %%%%%%%%%
%%%%%%%%%%%% the original texts by Denghang %%%%%%%%%
\hide{
Moreover, the transitions $(q, \myvec{0}, q')$ in $\cC_i$ can be seen as $\varepsilon$-transitions, that is, we can replace each transition $(q, \myvec{0}, q')$ with $(q, \varepsilon, q')$, without affecting the set $\{ \sum\limits_{\myvec{v} \in V_i} m_{\myvec{v}} \myvec{v} \mid \myvec{m} \in parikh(\cC_i)\}$. Therefore, we first do these replacements and apply the $\varepsilon$-transition removal procedure. Then we utilize the determinization and minimization algorithm for NFA to reduce the size of $\cC_i$.

It turns out that to solve the non-emptiness of CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ w.r.t. $\varphi$. We can ignore the characters in $\cB_1,\cdots, \cB_n$ because there is only one CEFA for each string variable, and $\varphi$ only cares about the values of registers.

@@

Moreover, the transitions $(q, \myvec{0}, q')$ in $\cC_i$ can be seen as $\varepsilon$-transitions, that is, we can replace each transition $(q, \myvec{0}, q')$ with $(q, \varepsilon, q')$, without affecting the set $\{ \sum\limits_{\myvec{v} \in V_i} m_{\myvec{v}} \myvec{v} \mid \myvec{m} \in parikh(\cC_i)\}$. Therefore, we first do these replacements and apply the $\varepsilon$-transition removal procedure. Then we utilize the determinization and minimization algorithm for NFA to reduce the size of $\cC_i$.

@@

In summary, we obtain an algorithm for solving the non-emptiness of $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ w.r.t. $\varphi$, as illustrated in Figure~\ref{xxx}.

@@

As shown in the Algorithm \ref{alg:underApprox}, $\varphi_{under}$ is the under-approximated LIA formula which is initialized to the input LIA formula $\psi$ at line 2. $\varphi_\aut$ is the under-approximated LIA formula of each automaton $\aut$ which is initialized to $false$ at line 3. $lastStates$ maps each automaton $\aut$ to the set consisting of pairs of state and corresponding registers' value in the last enumeration. $lastStates(\aut)$ is initialized to the pair of the initial state and zero values for each automaton $\aut$ at line 4. $length$ is the current bound of lengths of string variables which is initialized to $0$ at line 5. The main loop at lines 6-20 iteratively tries possible solutions from low to high length bound. For each iteration, we generate the LIA formula respectively for each automaton. To each CEFA $\aut$, we get pairs of last state and corresponding registers' values from $lastStates$ at line 8. If the last state is accepted at line 9 in each pair, we add the LIA formula of the possible registers' values to $\varphi_\aut$ as a conjunct at line 10. After that, we update the set of pairs in $lastStates$ by removing the probed pair at line 12 and adding new pairs of the next state and corresponding registers' values at line 13. Finally, we combine the incremental LIA formula of each automaton at line 15. If the LIA formula $\varphi\wedge \theta$ is satisfiable at line 17, i.e., we find a solution with all strings having length less than or equal to $length$, we return \emph{true} at line 18. Otherwise, the next iteration would be invoked with a larger length. If we can not find a solution within the max string length bound, we return \emph{false} at line 22.

@@

  Consider the automaton $\aut$ in Fig. \ref{fig:overview:product:reduced} again without the LIA constraint $r_0 > 120$, i.e., $\theta$ changes to $1\leq r_1\leq 60\wedge 1\leq r_2 \leq 60\wedge 0\leq r_3 \leq 60 $. The solution of the Parikh image of it leads to a complete but inefficient result since the quantifier-free LIA formulas are solved in exponential time. Under-approximation can accelerate the solving process. In the first iteration of under-approximation, the pairs in the $lastStates(\aut)$ is the initial pair $(q_0', (0,0,0,0))$ and $\varphi_\aut$ maintains false at line 9 because $q_0'$ is not accepted. Then we update $lastStates(\aut)$ by removing the pair $(q_0', (0,0,0,0))$ and adding pairs $(q_0', (1,1,0,0)), (q_1', (1,0,1,0)), (q_2', (1,0,0,0))$. In the second iteration, $\varphi_\aut$ changes to $r_0=1\vee r_1=0\vee r_2=0\vee r_3 = 0$ because $q_2'$ is accepted. $\varphi_{under}$ is always equal to $\varphi_\aut$ since there is only one automaton. We solve the formula $\theta\wedge \varphi_{under}$ and find it is unsatisfiable. In the third iteration, the pairs in the $lastStates(\aut)$ with accepting state should be $(q_0', (2,2, 0, 0)), (q_1', (2,1,1,0))$, $(q_1', (2,0,2,0))$, $(q_2', (2,1,0,0)), (q_2', (2,0,1,0)),$ $(q_2', (2,0, 0, 0))$. The corresponding registers' values for the accepted state $q_2'$ are $(2,1,0,0), (2, 0, 1, 0), (2, 0, 0, 0)$, in which the value of $r_1$ and $r_2$ can not be both larger than 1 which is required by $\theta$. So the formula $\theta\wedge \varphi_{under}$ is still unsatisfiable. Finally, We obtain a pair $(q_2', (3,1,1,0))$ from the pair $(q_1', (2,1,1,0))$ and the transition $q_1'\xrightarrow[(1,0,0,0)]{} q_2'$, whose corresponding registers' values is a solution $(3,1,1,0)$, i.e., $r_0=3, r_1=1, r_2=1, r_3=0$.

### 2023-08-02 2023-1-19/sections/cefa.tex
grammerly check

In this section, we define the model of cost-enriched finite state automata (CEFA), which was introduced in \cite{atva2020} and will be used to solve the satisfiability problem of RECL later on. 
%
Intuitively, CEFA add write-only cost registers to finite state automata. Here by ``write-only'', we mean that the cost registers can only be written/updated, but cannot be read, that is, cannot be used to guard the transitions.

@@

We would like to remark that the aforementioned definition of CEFA is slightly different from that in \cite{atva2020}, where CEFA did not include accepting conditions.
Moreover, the accepting conditions $\alpha$ in CEFA are defined in a \emph{global} fashion in the sense that the accepting condition does not distinguish final states. This technical choice is made so that the determinization and minimization of NFA can be utilized to reduce the size of CEFA in the next section.

%
%and $\alpha(q_n)[\myvec{\myvec{v'}/\myvec{r}]$ is true, where $\myvec{v'} = \sum \limits_{j \in [n]} \myvec{v_j}}$. Note here \denghang{that} we assume that the initial values of all registers are zero and $\sum \limits_{j \in [n]} \myvec{v_j}$ is the vector of register values after all the transitions in the run are executed. A string $w$ is \emph{accepted} by $\aut$ if there is an accepting run of $\aut$ on $w$. The language defined by $\aut$, denoted by $\Lang(\aut)$, is the set of strings that are accepted by $\aut$.

@@

In the sequel, we define three operations on CEFA, namely, union, intersection, and concatenation. These three operations will be used for solving RECL constraints in the next section.

@@

The \emph{union} of $\aut_1$ and $\aut_2$, denoted by $\aut_1 \cup \aut_2$, is defined in a slightly more involved manner than NFA, as a result of the technical choice that the accepting conditions of CEFA do not distinguish final states. Two fresh auxiliary registers, say $r'_1,r'_2 \not \in R_1 \cup R_2$, are introduced so that the accepting condition knows whether a run is from $\aut_1$ or from $\aut_2$. Specifically, $\aut_1 \cup \aut_2 = (R', Q', \Sigma, \delta', I', F', \alpha')$, where 
\begin{itemize}
\item $R' = R_1 \cup R_2 \cup \{r'_1, r'_2\}$, $Q' = Q_1 \cup Q_2 \cup \{q'_0\}$ with $q'_0 \not \in Q_1 \cup Q_2$, $I' = \{q'_0\}$, 
%
\item $\delta'$ is the union of $\{(q'_0, a, q'_1, (\myvec{v_1}, \myvec{0}, 1,0)) \mid \exists q_1 \in I_1.\ (q_1, a, q'_1, \myvec{v_1}) \in \delta_1\}$, $\{(q'_0, a, q'_2, (\myvec{0}, \myvec{v_2}, 0, 1)) \mid \exists q_2 \in I_2.\ (q_2, a, q'_2, \myvec{v_2}) \in \delta_2\}$, $\{(q_1, a, q'_1, (\myvec{v_1}, \myvec{0}, 0, 0)) \mid (q_1, a, q'_1, \myvec{v_1}) \in \delta_1\}$, and $\{(q_2, a, q'_2, (\myvec{0}, \myvec{v_2}, 0, 0)) \mid (q_2, a, q'_2, \myvec{v_2}) \in \delta_2\}$, where $(\myvec{v_1}, \myvec{0}, 1,0)$ is a vector that updates $R_1$ by $\myvec{v_1}$, updates $R_2$ by $\myvec{0}$, and updates $r'_1,r'_2$ by $1,0$ respectively. Similarly for $(\myvec{0}, \myvec{v_2}, 0, 1)$, and so on.

>>>>

In this section, we define the cost-enriched finite state automata (CEFA), which was introduced in \cite{atva2020} and will be used to solve the satisfiability problem of RECL later on. 
%
Intuitively, CEFA adds write-only cost registers to finite state automata. Here by ``write-only'', we mean that the cost registers can only be written/updated but cannot be read; that is, they cannot be used to guard the transitions.

@@

We want to remark that the definition above of CEFA is slightly different from that in \cite{atva2020}, where CEFA did not include accepting conditions.
Moreover, the accepting conditions $\alpha$ in CEFA are defined in a \emph{global} fashion because the accepting condition does not distinguish final states. This technical choice is made so that the determinization and minimization of NFA can be utilized to reduce the size of CEFA in the next section.

%
%and $\alpha(q_n)[\myvec{\myvec{v'}/\myvec{r}]$ is true, where $\myvec{v'} = \sum \limits_{j \in [n]} \myvec{v_j}}$. Note here \denghang{that} we assume that the initial values of all registers are zero and $\sum \limits_{j \in [n]} \myvec{v_j}$ is the vector of register values after all the transitions in the run are executed. A string $w$ is \emph{accepted} by $\aut$ if there is an accepting run of $\aut$ on $w$. The language defined by $\aut$, denoted by $\Lang(\aut)$, is the set of strings accepted by $\aut$.

@@

In the sequel, we define three CEFA operations: union, intersection, and concatenation. The following section will use these three operations to solve RECL constraints.

@@

The \emph{union} of $\aut_1$ and $\aut_2$, denoted by $\aut_1 \cup \aut_2$, is defined in a slightly more involved manner than NFA, as a result of the technical choice that the accepting conditions of CEFA do not distinguish final states. Two fresh auxiliary registers say $r'_1,r'_2 \not \in R_1 \cup R_2$, are introduced so that the accepting condition knows whether a run is from $\aut_1$ or $\aut_2$. Specifically, $\aut_1 \cup \aut_2 = (R', Q', \Sigma, \delta', I', F', \alpha')$, where 
\begin{itemize}
\item $R' = R_1 \cup R_2 \cup \{r'_1, r'_2\}$, $Q' = Q_1 \cup Q_2 \cup \{q'_0\}$ with $q'_0 \not \in Q_1 \cup Q_2$, $I' = \{q'_0\}$, 
%
\item $\delta'$ is the union of $\{(q'_0, a, q'_1, (\myvec{v_1}, \myvec{0}, 1,0)) \mid \exists q_1 \in I_1.\ (q_1, a, q'_1, \myvec{v_1}) \in \delta_1\}$, $\{(q'_0, a, q'_2, (\myvec{0}, \myvec{v_2}, 0, 1)) \mid \exists q_2 \in I_2.\ (q_2, a, q'_2, \myvec{v_2}) \in \delta_2\}$, $\{(q_1, a, q'_1, (\myvec{v_1}, \myvec{0}, 0, 0)) \mid (q_1, a, q'_1, \myvec{v_1}) \in \delta_1\}$, and $\{(q_2, a, q'_2, (\myvec{0}, \myvec{v_2}, 0, 0)) \mid (q_2, a, q'_2, \myvec{v_2}) \in \delta_2\}$, where $(\myvec{v_1}, \myvec{0}, 1,0)$ is a vector that updates $R_1$ by $\myvec{v_1}$, updates $R_2$ by $\myvec{0}$, and updates $r'_1,r'_2$ by $1,0$ respectively. Similarly for $(\myvec{0}, \myvec{v_2}, 0, 1)$, and so on.
